PAPER DRAFT

### 2021-11-29 Papers/Polymethylation_scores/06_discussion.Rmd
comments from erin

In the longitudinal Fragile Families and Child Wellbeing birth cohort, we observed that prenatal maternal smoking was associated with DNA methylation in children's saliva samples from ages 9 and 15. Prenatal maternal smoking was associated with polymethylation scores across strata of child age and ancestry. Global methylation and epigenetic clocks were not associated with maternal smoking exposure demonstrating a specific and persistent signal of prenatal smoke exposure in DNA methylation. Polymethylation scores had reasonable accuracy for classifying prenatal maternal smoking. This accuracy was more consistent across time than when using individual *a priori* CpG sites.

Our findings are consistent with the previous literature on associations between prenatal maternal smoking and DNA methylation. We replicated top hits from previous epigenome wide association studies of DNA methylation in cord and peripheral blood[@Joubert:2016ue; @wiklund2019]. There are no previous studies of prenatal maternal smoking and saliva DNA methylation. However, previous work has shown that the majority of CpG sites are similarly methylated in blood and saliva (reviewed in @Langie2017).

We advance the literature on prenatal maternal smoking and DNA methylation by evaluating the persistence of the association as children age and its portability across tissue and ancestry. Polymethylation scores built using coefficients from meta-analysis of cord blood DNA methylation from primarily European-ancestry newborns were still associated with prenatal maternal smoking in our independent saliva samples from a diverse cohort at ages 9 and 15 [@Joubert:2016ue]. The portability of other risk scores, such as polygenic risk scores, across ancestries is a complex research area [@Martin2019] and evaluating the portability of epigenetic summary measures has been identified as a key area for evaluation [@Hüls2020]. In this case, the polymethylation score for prenatal maternal smoking appears to be portable across ancestry groups. Effect estimates were consistent across ancestries at age 9, although Hispanic and European genetic ancestry samples had lower effect estimates at age 15. This could reflect higher unreported smoking initiation in European and Hispanic ancestry teens in the United States than African ancestry teens. Genetic ancestry correlates with race, and White and Latino teens have much higher rates of teen smoking and earlier ages at initiation than Black children [@El-Toukhy2016]. Though we excluded children who reported own-smoking from our analytic sample, under-reporting of own-smoking could influence DNA methylation at ages 9 and 15, creating outcome misclassification. This could also explain the observed attenuation the effect on AHRR:cg05575921 methylation by age 15 in our sample, as DNA methylation at this probe is known to vary by personal cigarette smoking[@Dawes2019; @Philibert2020].

@@

Our analysis contributes to the development of an accurate methylation biomarker for prenatal maternal smoking by evaluating the impact of specific methodological choices on accuracy of classification. DNA methylation biomarkers may be especially susceptible to confounding by subject age at methylation measurement and cell-type proportion [@Hüls2020]. Accuracy of polymethylation score classification was similar when using coefficients from cord blood vs peripheral blood samples from older children. Accuracy was not improved by using two time-points of methylation measurement. Accuracy was improved when polymethylation scores used coefficients which incorporated cell-type control. Our analysis therefore suggests that cell-type control when both generating and applying coefficients for polymethylation scores may positively influence accuracy, although more work comparing coefficients from different populations is still needed.

Our analysis suggests that polymethylation scores may be more accurate than using single CpG sites as biomarkers. AHRR:cg05575921 is a consistent marker of prenatal maternal smoke exposure in meta-analyses of newborn cord blood[@Joubert:2016ue]. DNA methylation at AHRR:cg05575291 can accurately classify own--smoking behavior in both blood (area under the curve 0.995) and saliva (area under the curve 0.971) [@Dawes2019; @Philibert2020]. However, salivary AHRR:cg05575921 was not a persistent marker of prenatal maternal smoking in an analysis of middle-aged adult women [@Richmond:2018tu]. In our sample, salivary AHRR:CpG05575921 methylation was less accurate at predicting prenatal maternal smoke exposure than polymethylation scores. The accuracy of AHRR:CpG05575921 and other single CpG site biomarkers may be influenced by time-since-exposure and new environmental exposures. Incorporating information across multiple sites of DNA methylation may yield a biomarker more robust to these influences.

Our analysis is not without its limitations. We used maternal self report of prenatal maternal smoker as serum cotinine levels were not available. Due to social desirability bias, this could result in exposure misclassification. We would expect this to bias our results towards the null. In any analysis of a prenatal exposure and postnatal outcome, there is the possibility of selection bias into the cohort due to live birth bias. Selection bias is also possible due to loss-to-follow-up between birth and age 9. Additionally, while we controlled for postnatal secondhand smoke exposure and excluded children who reported any own-smoking, we cannot exclude the possibility of residual confounding in this observational cohort.

However, our analysis also has several strengths. We analyzed samples from a large sample of diverse participants underrepresented in genetic and epigenetic research [@Sirugo2019; @Bentley2017]. While our exposure measurement of prenatal maternal smoking was self reported, it was assessed prospectively and preceded outcome measurements. We analyzed repeated measures of DNA methylation with reproducible array measures conducted in a single batch. We tested associations between prenatal maternal smoking and multiple DNA summary measures to evaluate the specificity of the polymethylation scores. In sensitivity analyses, we adjusted for other prenatal exposures and postnatal smoke exposure to examine the specificity of the biomarker to nature and timing of exposure.

>>>>

In the longitudinal Fragile Families and Child Wellbeing birth cohort, we observed that prenatal maternal smoking was associated with several characterizations of DNA methylation in children's saliva samples from ages 9 and 15. Prenatal maternal smoking was associated with polymethylation scores across strata of both child age and genetic ancestry. Global methylation and epigenetic clocks were not associated with maternal smoking exposure. Polymethylation scores had reasonable accuracy for classifying prenatal maternal smoking (AUC \> 0.7). Classification when using polymethylation scores for prenatal maternal smoking was better than when using a single *a priori* CpG site, cg05575921 in the *AHRR* gene.

Our findings are consistent with the previous literature on associations between prenatal maternal smoking and DNA methylation. We replicated the top hit from a previous epigenome wide association analysis [@Joubert:2016ue]. Further, we found evidence of association in several additional hits from epigenome wide association studies of DNA methylation in cord and peripheral blood [@Joubert:2016ue; @wiklund2019]. There are no previous studies of prenatal maternal smoking and saliva DNA methylation. However, previous work has shown that the majority of CpG sites are similarly methylated in blood and saliva (reviewed in @Langie2017).

We advance prenatal smoking - DNA methylation literature by evaluating the persistence of the association between prenatal maternal smoking and DNA methylation as children age and its portability across tissue and ancestry. Polymethylation scores built using coefficients from meta-analysis of cord blood DNA methylation from primarily European-ancestry newborns were still associated with prenatal maternal smoking in our independent saliva samples from a diverse cohort at ages 9 and 15 [@Joubert:2016ue]. The portability of other risk scores, such as polygenic risk scores, across genetic ancestries is a complex research area [@Martin2019] and evaluating the portability of epigenetic summary measures has been identified as a key area for evaluation [@Hüls2020]. In this case, the polymethylation score for prenatal maternal smoking appears to be portable across genetic ancestry groups. Effect estimates were consistent across ancestries at age 9, although Hispanic and European genetic ancestry samples had lower, non-significant effect estimates at age 15. The decreased precision may be the result of the reduced sample size of the HIspanic and European genetic ancestry samples. The reduction in effect estimate magnitude could reflect higher unreported smoking initiation in European and Hispanic ancestry teens in the United States than African ancestry teens. Genetic ancestry correlates with race, and White and Latino teens have much higher rates of teen smoking and earlier ages at initiation than Black children [@El-Toukhy2016]. Though we excluded children who reported own-smoking from our analytic sample, under-reporting of own-smoking could influence DNA methylation at ages 9 and 15, creating outcome misclassification. This could also explain the observed attenuation of the effect on *AHRR*:cg05575921 methylation by age 15 in our sample, as DNA methylation at this probe is known to vary by personal cigarette smoking[@Dawes2019; @Philibert2020].

@@

Our analysis contributes to the development of an accurate methylation biomarker for prenatal maternal smoking by evaluating the impact of specific methodological choices on accuracy of classification. DNA methylation biomarkers may be especially susceptible to confounding by subject age at methylation measurement and cell-type proportion [@Hüls2020]. Prenatal maternal smoke exposure classification accuracy of polymethylation score was similar when using coefficients from cord blood vs peripheral blood samples from older children. Classification accuracy was not improved by using two time-points of methylation measurement. Classification accuracy was improved when polymethylation scores used coefficients which incorporated cell-type control. Our analysis therefore suggests that cell-type control when both generating (in the epigenome wide association study) and applying coefficients for polymethylation scores may positively influence outcome classification accuracy, although more work comparing coefficients from different populations is still needed.

Our analysis suggests that polymethylation scores may be more accurate than using single CpG sites as biomarkers. The site cg05575921 in the *AHRR* gene is a consistent marker of prenatal maternal smoke exposure in meta-analyses of newborn cord blood[@Joubert:2016ue]. DNA methylation at cg05575291 in the *AHRR* gene can accurately classify own--smoking behavior in both blood (area under the curve 0.995) and saliva (area under the curve 0.971) [@Dawes2019; @Philibert2020]. However, salivary AHRR:cg05575921 was not a persistent marker of prenatal maternal smoking in an analysis of middle-aged adult women [@Richmond:2018tu]. In our sample, salivary AHRR:CpG05575921 methylation was less accurate at predicting prenatal maternal smoke exposure than polymethylation scores. The accuracy of AHRR:CpG05575921 and other single CpG site biomarkers may be influenced by time-since-exposure and new environmental exposures. Incorporating information across multiple sites of DNA methylation may yield a biomarker more robust to these influences.

Our analysis is not without its limitations. We used maternal self-report of prenatal maternal smoke exposure, as serum cotinine levels were not available. Due to social desirability bias, this could result in exposure misclassification. We would expect this to bias our results towards the null. In any analysis of a prenatal exposure and postnatal outcome, there is the possibility of selection bias into the cohort due to live birth bias. Selection bias is also possible due to loss-to-follow-up between birth and age 15. Additionally, while we controlled for postnatal secondhand smoke exposure and excluded children who reported any own-smoking, we cannot exclude the possibility of residual confounding in this observational cohort.

However, our analysis also has several strengths. We analyzed samples from a large cohort of diverse participants underrepresented in genetic and epigenetic research [@Sirugo2019; @Bentley2017]. While our exposure measurement of prenatal maternal smoking was self-reported, it was assessed prospectively and preceded outcome measurements. We analyzed repeated measures of DNA methylation with reproducible array measures conducted in a single batch. We tested associations between prenatal maternal smoking and multiple DNA summary measures to evaluate the specificity of the polymethylation scores. In sensitivity analyses, we adjusted for other prenatal exposures and postnatal smoke exposure to examine the specificity of the biomarker to nature and timing of exposure.

### 2021-11-29 Papers/Polymethylation_scores/07_conclusions.Rmd
comments from erin

In a large, prospective study of diverse participants, we showed that DNA methylation in children's saliva had strong associations with and reasonable accuracy for prenatal maternal smoke exposure. Further, we demonstrated that polymethylation scores could be applied as a biomarker of prenatal maternal smoke exposure across ancestry-groups, an important consideration for the equitable biomarker development. The development and application of biomarker for prenatal maternal smoke exposure has important implications for epidemiological research and clinical practice. Given the difficulty of measuring prenatal maternal smoke exposure, such a biomarker could allow for confounder control in research areas where such control is currently impossible. Prenatal maternal smoke exposure is prevalent and has negative health consequences, thus an exposure biomarker could be used to provide support and health interventions for children.

>>>>

In a large, prospective study of diverse participants, we showed that DNA methylation in children's saliva had strong associations with and reasonable classification accuracy for prenatal maternal smoke exposure. Further, we demonstrated that polymethylation scores could be applied as a biomarker of prenatal maternal smoke exposure across genetic ancestry groups, an important consideration for the equitable biomarker development. The development and application of biomarkers for prenatal maternal smoke exposure has important implications for epidemiological research and clinical practice. Given the difficulty of measuring prenatal maternal smoke exposure, such a biomarker could allow for confounder control in research areas where such control is currently impossible. Prenatal maternal smoke exposure is prevalent and has negative health consequences, thus an exposure biomarker could be used to provide support and health interventions for children.

### 2021-11-29 Papers/Polymethylation_scores/08_methods.Rmd
comments from erin

Child genetic ancestry was calculated. Principal components of child genetic ancestry were calculated from genetic data (CHIP?) from child saliva samples.

@@

DNA methylation image data were processed in R statistical software (3.5) using the minfi package [@Aryee2014]. The red and green image pairs (n=1811) were read into R and the minfi *preprocessNoob* function was used to normalize dye bias and apply background correction. Further quality control was applied using the ewastools packages [@Heiss2018]. We dropped samples with \>10% of sites have detection p-value \>0.01 (n=43), sex discordance between DNA methylation predicted sex and recorded sex (n=20), and abnormal sex chromosome intensity (n=3). CpG sites were removed if they had detection p-value \>0.01 in 5% of samples (n=26,830) or were identified as cross-reactive (n=27,782) (Figure 1) [@Chen2013]. Relative proportions of immune and epithelial cell types were estimated from DNA methylation measures using a childhood saliva reference panel [@Middleton2021].

We created polymethylation scores for prenatal maternal smoke exposure. From an independent meta-analysis of prenatal smoke exposure and newborn DNA methylation, we extracted the regression coefficients of 6,074 CpG sites associated with prenatal maternal smoking at a false discovery ratio corrected $P$ value \<0.05 [@Joubert:2016ue]. We mean-centered the DNA methylation beta values in our study, weighted them by the independent regression coefficients and took the sum. We calculated polymethylation scores using regression coefficients from 4 different regressions. For our main analysis we used coefficients from a regression of sustained smoking exposure and DNA methylation in newborn cord blood with cell-type control. As sensitivity analyses, we used re sustained smoking exposure and DNA methylation in newborn cord blood without cell-type control, sustained smoking exposure and DNA methylation in peripheral blood from older children, any smoking exposure and DNA methylation in newborn cord blood) [@Joubert:2016ue].

>>>>

Child genetic ancestry was calculated. Grants R01 HD36916, R01 HD073352, and R01 HD076592 provided support for the collection, assay, and analysis of the genetic data in the Fragile Families cohort. Principal components (PC) of child genetic ancestry were calculated from genetic data using Illumina PsychChip_v1-1 with child saliva samples. Genetic ancestry was assigned by comparing PC loadings to 1000 Genomes super population clusters. Samples with PC1 \> 0.018 and PC2 \> -0.0075 were assigned to European ancestry. Samples with PC1 \< -0.005 and PC2 \> 0.007+0.75(PC1) were assigned to African ancestry. Samples with PC1 \> 0.018 and -0.055 \< PC2 \< 0.025 were assigned to Hispanic ancestry.

@@

DNA methylation image data were processed in R statistical software (3.5) using the minfi package [@Aryee2014]. The red and green image pairs (n=1811) were read into R and the minfi *preprocessNoob* function was used to normalize dye bias and apply background correction. Further quality control was applied using the ewastools packages [@Heiss2018]. We dropped samples with \>10% of sites have detection p-value \>0.01 (n=43), sex discordance between DNA methylation predicted sex and recorded sex (n=20), and abnormal sex chromosome intensity (n=3). CpG sites were removed if they had detection p-value \>0.01 in 5% of samples (n=26,830) or were identified as cross-reactive (n=27,782) (Figure 1) [@Chen2013]. We used the November 2021 data freeze of the Fragile Families and Child Wellbeing DNA methylation data. Relative proportions of immune and epithelial cell types were estimated from DNA methylation measures using a childhood saliva reference panel [@Middleton2021].

We created polymethylation scores for prenatal maternal smoke exposure. From an independent meta-analysis of prenatal smoke exposure and newborn DNA methylation, we extracted the regression coefficients of 6,074 CpG sites associated with prenatal maternal smoking at a false discovery ratio corrected $P$ value \<0.05 [@Joubert:2016ue]. We mean-centered the DNA methylation beta values in our study, weighted them by the independent regression coefficients and took the sum. We calculated polymethylation scores using regression coefficients from 4 different regressions. For our main analysis we used coefficients from a regression of sustained smoking exposure and DNA methylation in newborn cord blood with cell-type control. As sensitivity analyses, we used sustained smoking exposure and DNA methylation in newborn cord blood without cell-type control, sustained smoking exposure and DNA methylation in peripheral blood from older children, any smoking exposure and DNA methylation in newborn cord blood) [@Joubert:2016ue].

### 2021-11-29 Papers/Polymethylation_scores/98_declarations.Rmd
comments from erin

The authors have no competing interests to declare. check here

@@

This research was made possible through several grants (add). FB was supported by the National Institutes of Health National Institute of Dental and Craniofacial Research (F31DE029992).

@@

FB performed the analysis with contributions from JD, JF, and EW. FB and KB wrote the paper with all authors contributing to revisions. DN, CM, KM, EW and FB all contributed to the design and conception of the study.

>>>>

The authors have no competing interests to declare.

@@

This research was made possible through several grants (Fragile Families Core Data Collection: R01 HD036916, R25 AG053227, genetic data processing R01 HD36916, R01 HD073352, and R01 HD076592). FB was supported by the National Institutes of Health National Institute of Dental and Craniofacial Research (F31 DE029992). EB was supported by (R01 AG055406, R01 AG067592, R01 MD011716).

@@

FB performed the analysis with contributions from JD, JF, and EW. FB and KB wrote the paper with all authors contributing to revisions. CM, DN and LS contributed to the design and processing of the DNA methylation data subcohort. DN, CM, KM, EW, and FB all contributed to the design and conception of the analytic plan.  All authors reviewed and revised the manuscript.
